Crossing Origins by Crossing Formats

### 2012-11-29 main.tex
Major update to section "Vulnerabilities and attacks"

In this section we give concrete examples of the attacks described in the
previous section, using the PDF format as the running example.

\subsection{Concept}

\subsection{The PDF format}

% Why PDF format?

The PDF format is a widely used, text based format with powerful capabilities.
It is capable of displaying text, rendering graphics, animation and dynamic
content. It is a container format which allows for embedding of other files
and formats. This and the fact that it is designed to be robust against errors,
provides us with a format that is powerful, flexible and ideal for creating
malicious polyglots.

% Several design choices

% - Error tolerant parsing
% - Execute JavaScript
% - Embed flash files
% - Launch commands

Furthermore the PDF format is of particular interest to us because it can
execute JavaScript and send HTTP requests to the server from which the document
originated.

\subsubsection{PDF communication channels}

For communication with the containing page, one can either use
hostContainer.postMessage(info) or app.launchURL('\#'+info). The latter has the
advantage that it works in Firefox which seems to not handle the postMessage
very well.

\subsection{The object-tag}

%The object-tag allows the developer to specify which format 
%the embedded content should be interpreted as.

The object-tag is used to embed content, which requires a browser plug-in
to run, in a web page. The plug-in is selected based on the MIME-type of the content, but
because the server delivering the content might not be able to determine its
format, the object tag allows a developer to set the type-attribute to guide the
browser in which plug-in to run. When the type attribute is used, the
corresponding plug-in is run regardless of which MIME-type the server returns
and it is up to the plug-in to respond to and handle the content it is served.

As it turns out, various browser plug-ins deal with the situation differently. The Chrome browser
built-in PDF reader refuses to render content that was served with an inappropriate 
MIME-type. The Java plug-in used to ignore the MIME-type, but after several 
vulnerabilities were disclosed, this behavior was changed.

@@

@@

\begin{lstlisting}[label=lst:minimal-pdf, caption=HTML / E4X Polyglot]
%PDF-
trailer<</Root<</Pages<<>>>>>>
\end{lstlisting}

>>>>

In this section we give concrete examples of the attacks described in 
Section~\ref{sec:cross}, using the PDF format as the running example. We begin
by detailing how the design decision made in the PDF-standard make it ideal 
for creating malicious polyglots.

%\subsection{Concept}

\subsection{Portable Document Format}
\label{sec:format}
% Why PDF format?

The \emph{Portable Document Format} (PDF) is a widely used document format 
capable of displaying text, rendering graphics, scripting, animation and other 
dynamic content. It is a container format in the sense that it allows 
embedding of files and resources.

According to the PDF specification \todo{cite specification} a PDF file is composed of
a header, several objects, a cross-reference section and a trailer. Listing~\ref{lst:sample-pdf} 
shows a sample of how a PDF file is structured and its elements.

% header
\subsubsection{Header}
The header consists of the string "\%PDF-" followed by a version number. 
The version is denoted by a 
major and a minor version number of the form "\emph{M.m}". Because the version can 
be specified elsewhere, the version number is not required to be part of the 
header.

% objects
\subsubsection{Objects}
Objects can be direct or indirect, the difference being that indirect objects 
has labels which are used for referring to the object from another object. Object 
labels are numbered and begin with the string "\emph{N n} obj", where "N" is the object number and "n" is the 
revision number. Similarly object references are of the form "\emph{N n} R".
The label is optionally ended with the keyword "endobj".

There are eight basic types of basic objects; booleans, integers, strings, names, 
arrays, dictionaries, streams and the null object. For the intents and purposes 
of this article, we will focus on the string, dictionary, name and stream objects.

% string objects
\paragraph{String objects}
There are two types of strings; literal and hexadecimal. Literal strings are 
enclosed by the "(" and ")" characters. Any character can occur in a literal 
string, even parentheses if they are balanced, e.g. a matching closing 
parenthesis for every opened parenthesis. In hexadecimal strings 
each character is represented by its corresponding hexadecimal value, enclosed 
by the "$<$" and "$>$" characters.

% dictionary objects
\paragraph{Dictionary objects}
Dictionary objects are a name-value map delimited by the "$<<$" and "$>>$" 
tokens. The names are name objects and the values are objects of any type.
% name objects
Name objects begin with the "/" character, followed by a string of non-whitespace 
characters. Each dictionary has a type, either specified by the "/Type" name or inferred 
from the context in which it occurs. The type declares which kind of element the 
dictionary is describing, e.g. a page or an annotation. Dictionaries form the structure of the document by connecting
objects through references, e.g. relating a page to its contents. %At least one 
%dictionary is required to build a PDF document. 
% actions?
A special type of dictionary describes \emph{actions}. Actions are triggered when a
certain event occur, such as a file is opened or a page is displayed, and the action 
dictionary specify how it should be handled. Actions can be used to, among other 
things, go to a specific page, play a sound, execute JavaScript or launch a command.

% stream objects
\paragraph{Stream objects}
A stream is an unlimited sequence of bytes. A stream object consists of a 
dictionary, describing the stream, and the associated stream delimited by the 
"stream" and "endstream" keywords. PDF supports encoding of streams, in which
case the dictionary describe which filters are required to decode the stream.

% cross-reference
\subsubsection{Cross-reference}
The cross-reference section is a record of the location of indirect objects 
within a file. The location is specified as the byte offset from the beginning 
of the file. The cross-reference section is opened with the "xref" keyword,
followed by one record for each indirect object. The cross-reference section 
will be reconstructed if missing and can therefore be omitted.

% trailer
\subsubsection{Trailer}
The trailer is composed of a trailer dictionary, a pointer to the cross-reference 
section and an end-of-file marker. The trailer dictionary is introduced by the 
"trailer" keyword. It contains among other things a reference to the root of the 
document structure. The "startxref" keyword is followed by the number of bytes from 
the beginning of the file to the first cross-reference section. As with the 
cross-reference section, it can be omitted. The string "\%\%EOF" marks the 
end-of-file, but can be omitted.

\begin{lstlisting}[label=lst:sample-pdf, caption=Sample PDF file]
%PDF-1.7
1 0 obj<</Type/Catalog/Pages<<>>>>
endobj
2 0 obj<</Length 13>>stream
Sample Author
endstream
endobj
xref
0 3
0000000000 65536 f
0000000009 00000 n
0000000057 00000 n
trailer<</Root 1 0 R
	/Info<</Author 2 0 R>>
>>
startxref
118
%%EOF
\end{lstlisting}

\subsection{PDF-based polyglots}
Several design choices in the PDF specification make the format particularly 
suitable for making polyglots. One such design choice is the error tolerant 
parser. This is in part motivated it being a container format, of which the 
implication is that a PDF file can, by design, contain foreign syntax that 
could interfere with the syntax of the file. Another motivation is that PDF 
files are designed to be both forward- and backward-compatible. Readers 
implementing an older version of the specification do not recognize new 
features and behave as if they were not present in the file. 
%
The implementation notes of the specification describes some exceptions to the
requirements of the specification, such as the header can occur 
anywhere within the first 1024 bytes of the file. This flexibility gives plenty of room for 
combining with syntax of another format.
%
The specification declares many components to be required in a
PDF file, but as can be seen in Section~\ref{sec:format}, in practice several 
components can be omitted. The code in Listing~\ref{lst:minimal-pdf} shows the 
minimal syntax required for a malformed, but valid PDF file.\todo{cite Julia Wolf?}
%incrementally 
%updated by appending updated versions of existing objects to 
%the end of the file. In fact, 
%and formats. This and the fact that it is designed to be robust against errors,
%provides us with a format that is powerful, flexible and ideal for creating
%malicious polyglots.

% - Execute JavaScript
% - Embed flash files
% - Launch commands
% - Network requests

Furthermore the PDF format is of particular interest to us because of its 
many capabilities. Some examples include executing JavaScript, launching commands, 
and issuing HTTP requests. Adobe Reader even includes a flash 
player to play embedded flash files on systems that do
not have a player installed. %Considering that it might be a conscious decision
%of the user to refuse a flash player,

\begin{lstlisting}[label=lst:minimal-pdf, caption=Minimal PDF file]
%PDF-
trailer<</Root<</Pages<<>>>>>>
\end{lstlisting}

\subsection{The object- and embed-tags}
\todo{consider renaming to "Embedded PDF documents", and describe the plug-in behaviour, origin of content, type handling, cookies?}
%The object-tag allows the developer to specify which format 
%the embedded content should be interpreted as.

\todo{consider moving to section "Crossing origins"}
The object- and embed-tags are used to embed content, which requires a browser plug-in
to run, in a web page. The plug-in is selected based on the MIME-type of the content, but
because the server delivering the content might not be able to determine its
format, the object tag allows a developer to set the type-attribute to guide the
browser in which plug-in to run. When the type attribute is used, the
corresponding plug-in is run regardless of which MIME-type the server returns
and it is up to the plug-in to respond to and handle the content it is served.

\todo{consider revising}
As it turns out, various browser plug-ins deal with the situation differently. The Chrome browser
built-in PDF reader refuses to render content that was served with an inappropriate 
MIME-type. The Java plug-in used to ignore the MIME-type, but after several 
vulnerabilities were disclosed, this behavior was changed.

@@

\subsection{PDF communication primitives}

% Cross-origin leaks
To leak information across origins, a communication channel is 
required between the attacker and the target. PDF provides the necessary means 
to mediate this information leakage. The flow of information can be divided into 
two steps, extracting information from the target and delivering it to the attacker.
Each requiring its own communication primitives.

When embedded in a web page, many of the PDF communication primitives utilize
the underlying browser to make requests by navigating the browser to the target 
URL. This will unload the current page in which the document is embedded. Such 
primitives can be used in Cross-Site Request Forgery (CSRF) attacks, but would 
defeat the purpose of requesting information as the document cannot process the 
response.

\subsubsection{The target}
There are two primitives for a PDF to communicate with the server it originated from, 
that allow it to read the response; from JavaScript using XML external entities 
or from an embedded flash file.

\paragraph{XML}
The PDF JavaScript API includes a method to parse an XML document, called 
XMLData.parse. The XML document being processed may rely on external entities, 
which the XML parser will request and include in the document. The request is 
bound to the origin of the document. As the response 
is included in the document, it has to be valid, well-formed XML. This puts a 
restriction on the content that can be requested using this method. Considering 
that HTML pages are rarely well-formed this method may be too restricted.
Listing~\ref{lst:xml-pdf} contains an example that uses XML external entities.

\paragraph{Embedded Flash}
The embedded flash file inherits the origin of the document, thus it can 
request and read any document from the originating server. Unlike the XML
method, there is no restrictions on what content can be requested.

\begin{lstlisting}[label=lst:xml-pdf, caption=PDF using XML for communication]
%PDF-
1 0 obj<<>>stream
xml = '<!DOCTYPE foo[<!ENTITY x SYSTEM "'+
	URL+'">]><foo>&x;</foo>'
app.alert(XMLData.parse(xml).saveXML())
endstream
trailer<</Root<</Pages<<>>
    /OpenAction<</S/JavaScript/JS 1 0 R>>
    >>
>>
\end{lstlisting}

\subsubsection{The attacker}
There are a number of ways to communicate with the attacker, either directly 
via the embedding page or through the attackers server. While it is not To prevent the document from unloading

\paragraph{PostMessage}
The PDF plugin includes an API for communicating directly with the page in 
which it is embedded. The document can send messages to the page using 
the hostContainer.postMessage method.

\paragraph{Hash transport}

\paragraph{Embedded Flash}
Cross-origin communication in Flash is restricted by Cross-Origin Resource Sharing. 
As mentioned, the embedded Flash file inherits the origin of the document.
However, this is not a restriction, since an attacker can set a cross-origin policy 
that allow the request.

@@
